Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-05-30 method.tex
Tweaked table 1
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@769 ccdd05d5-56c1-461a-bf98-85f6876523b4
Fixed run-on
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@773 ccdd05d5-56c1-461a-bf98-85f6876523b4
Fix a few macros without following spaces.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@778 ccdd05d5-56c1-461a-bf98-85f6876523b4

Inspiral signals are parameterized by a set of intrinsic parameters
$\bar{\theta}$ that determine the amplitude and phase evolution of the observed
binary signals. For systems where the effects of spin can be ignored, the
intrinsic parameters are the component masses of the binary, $\bar{\theta} =
(m_1, m_2)$. Searches for gravitational waves from compact binary coalescence
typically employ matched filter banks, called template banks, that discretely
sample the possible intrinsic parameters~\cite{findchirppaper}.  The filters
for the waveforms $x(t,\bar{\theta})$, known as templates, are the waveforms
weighted by the inverse detector noise amplitude spectral density, so as to
whiten them. To construct a template bank, templates are chosen with discrete
signal parameters $\theta_1,\, \theta_2,\, \cdots$, $\theta_N$ to assure a
bounded loss of \SNR~\cite{Owen:1995tm,Owen:1998dk}. That is, any possible
signal within the search space will have a cross-correlation of $\geq0.97$ with
at least one template. Such a template bank is said to have a 97\% {\em
minimum match}. Data are filtered against each template to produce an \SNR
time-series. Local peak-finding across time and templates determines detection
candidates.
%
\footnote{There are two gravitational-wave polarizations, $+$ and $\times$. A
given detector will observe a combination of these polarizations that will
largely be degenerate with an overall unknown constant phase.  This can be
maximized over by filtering for quadrature phases and taking the magnitude of
the result. For simplicity we will ignore that aspect in this work, as it is
straightforward to generalize but not necessary for understanding any of the
points that will be made.}

@@

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
filter.  The time slices for this template consist of the $k$ intervals $(t_k,
t_{k-1}], \dots, (t_2, t_1], (t_1, t_0]$ sampled at frequencies
$f_\mathrm{k-1}, \dots, f_1, f_0$ where $f_0 \geq 2 f_\mathrm{ISCO}$, $t_0 =
t_\mathrm{ISCO}$, $f_{k-1} \geqslant 2 f_\mathrm{low}$, and $t_k \leqslant
t_\mathrm{low}$. An example time-slice design satisfying these constraints for
a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f_k$ is the sample
rate of the time slice, $(t_{k+1}, t_k]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$k^{\mathrm{th}}$ filter.}
\end{table}
%

@@

The \fir\ filter implementation has the advantage that it has no intrinsic
latency, whereas the \fft\ convolution has latency of $\fftblock-\tmpsamps$. 
%
%
%For example, for a $1.4 -
%1.4 \, M_\odot$ template with duration $\sim 1 \, \mathrm{ks}$, the
%\textsc{fft} convolution has a latency $\geq 2 \, \mathrm{ks}$.  
%
However, the \fir\ filter implementation has the disadvantage of much greater
overhead per sample than the \fft\ convolution.  For a $1\,\mathrm{ks}$
template sampled at $4096\,\mathrm{Hz}$, the \textsc{fir} implementation
requires about about $\tmpsamps / 8 \lg 2 \tmpsamps = 2.2 \times 10^4$ times
more operations per sample than the \fft implementation.  We will now consider
the computational cost of the \fir\ filter implementation described in the
previous sections.

It is convenient to express the computational cost of the entire filtering
procedure in floating point operations per second {\flops}.  The cost will be
the sum of the cost of the \fir\ filtering for the orthogonal filter in each
time slice plus the cost of reconstructing the original waveforms with matrix
operations and resampling.  Using the formulas in table \ref{table:flops} we
arrive at
%
\begin{equation}
%\mathrm{Computational\, cost} 
\mathrm{C.C.} \propto 2 \sum_k^{\numslices} (\slicessamps + \numtmps) \numsvdtmps f_k \nonumber
	r \sum_{k, f_k \neq f_{\fmax}}^{\numslices} \numtmps \resampsamps f_k 
\label{eq:TDcost}
\end{equation}
%
where the first sum is the total cost for filtering and reconstructing the
orthogonal filters $u_{jk}$.  The second sum is the cost of resampling the
reconstructed time-slice outputs to the original sample rate {\fmax}.  It
assumes a \fir\ filter with \resampsamps\ sample points.  Note that the cost
for resampling only occurs for the downsampled time slices. The resampling cost
largely depends on {\resampsamps}.  The computational cost of \eqref{eq:TDcost}
is dominated by the highest frequency terms in the sum.  Comparing just the
highest frequency term of \eqref{eq:TDcost} with \eqref{eq:SNRTD} shows \order{
(\slicesamps{\mathrm{max}} + \numtmps) \svdtmps{\mathrm{max}} \fmax} \flops\
versus \order{\numtmps \tmpsamps \fmax} {\flops}.  The full calculation for a
particular patch of parameter space is shown in table \ref{table:comp_cost}.
\hl{FIXME we need to actually say what went into this}

>>>>

Inspiral signals are parameterized by a set of intrinsic parameters
$\bar{\theta}$ that determine the amplitude and phase evolution of the observed
binary signals. For systems where the effects of spin can be ignored, the
intrinsic parameters are the component masses of the binary, $\bar{\theta} =
(m_1, m_2)$. Searches for gravitational waves from compact binary coalescence
typically employ matched filter banks, called template banks, that discretely
sample the possible intrinsic parameters~\cite{findchirppaper}.  The filters
for the waveforms $x(t,\bar{\theta})$, known as templates, are the waveforms
weighted by the inverse detector noise amplitude spectral density, so as to
whiten them. To construct a template bank, templates are chosen with discrete
signal parameters $\theta_1,\, \theta_2,\, \cdots$, $\theta_N$ to assure a
bounded loss of \SNR~\cite{Owen:1995tm,Owen:1998dk}. That is, any possible
signal within the search space will have a cross-correlation of $\geq0.97$ with
at least one template. Such a template bank is said to have a 97\% {\em
minimum match}. Data are filtered against each template to produce an \SNR{}
time-series. Local peak-finding across time and templates determines detection
candidates.
%
\footnote{There are two gravitational-wave polarizations, $+$ and $\times$. A
given detector will observe a combination of these polarizations that will
largely be degenerate with an overall unknown constant phase.  This can be
maximized over by filtering for quadrature phases and taking the magnitude of
the result. For simplicity we will ignore that aspect in this work, as it is
straightforward to generalize but not necessary for understanding any of the
points that will be made.}

@@

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
filter.  The time slices for this template consist of the $k$ intervals $(t_k,
t_{k-1}], \dots, (t_2, t_1], (t_1, t_0]$ sampled at frequencies
$f_\mathrm{k-1}, \dots, f_1, f_0$ where $f_0 \geq 2 f_\mathrm{ISCO}$, $t_0 =
t_\mathrm{ISCO}$, $f_{k-1} \geqslant 2 f_\mathrm{low}$, and $t_k \leqslant
t_\mathrm{low}$. An example time-slice design satisfying these constraints for
a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f_k$ is the sample
rate of the time slice, $(t_{k+1}, t_k]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$k^{\mathrm{th}}$ filter.}
\end{table}
%

@@

The \fir\ filter implementation has the advantage that it has no intrinsic
latency, whereas the \fft\ convolution has latency of $\fftblock-\tmpsamps$. 
%
%
%For example, for a $1.4 -
%1.4 \, M_\odot$ template with duration $\sim 1 \, \mathrm{ks}$, the
%\textsc{fft} convolution has a latency $\geq 2 \, \mathrm{ks}$.  
%
However, the \fir\ filter implementation has the disadvantage of much greater
overhead per sample than the \fft\ convolution.  For a $1\,\mathrm{ks}$
template sampled at $4096\,\mathrm{Hz}$, the \textsc{fir} implementation
requires about about $\tmpsamps / 8 \lg 2 \tmpsamps = 2.2 \times 10^4$ times
more operations per sample than the \fft{} implementation.  We will now consider
the computational cost of the \fir\ filter implementation described in the
previous sections.

It is convenient to express the computational cost of the entire filtering
procedure in floating point operations per second {\flops}.  The cost will be
the sum of the cost of the \fir\ filtering for the orthogonal filter in each
time slice plus the cost of reconstructing the original waveforms with matrix
operations and resampling.  Using the formulas in table \ref{table:flops} we
arrive at
%
\begin{equation}
%\mathrm{Computational\, cost} 
\mathrm{C.C.} \propto 2 \sum_k^{\numslices} (\slicessamps + \numtmps) \numsvdtmps f_k \nonumber
	r \sum_{k, f_k \neq f_{\fmax}}^{\numslices} \numtmps \resampsamps f_k 
\label{eq:TDcost}
\end{equation}
%
where the first sum is the total cost for filtering and reconstructing the
orthogonal filters $u_{jk}$.  The second sum is the cost of resampling the
reconstructed time-slice outputs to the original sample rate {\fmax}.  It
assumes a \fir\ filter with \resampsamps\ sample points.  Note that the cost
for resampling only occurs for the downsampled time slices. The resampling cost
largely depends on {\resampsamps}.  The computational cost of \eqref{eq:TDcost}
is dominated by the highest frequency terms in the sum.  Comparing just the
highest frequency term of \eqref{eq:TDcost} with \eqref{eq:SNRTD} shows \order{
(\slicesamps{\mathrm{max}} + \numtmps) \svdtmps{\mathrm{max}} \fmax} \flops{}
versus \order{\numtmps \tmpsamps \fmax} {\flops}.  The full calculation for a
particular patch of parameter space is shown in table \ref{table:comp_cost}.
\hl{FIXME we need to actually say what went into this}
